---
title: "Predict Epidemic Curves for COVID-19 with Generalized SEIR Modeling"
author: "Selcuk Korkmaz"
date: "22/06/2021"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    includes:
      in_header: header.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r current, include=FALSE}
currentDate <- as.Date(Sys.Date()-1, format = "%m/%d/%y")
```

## Implementation of genSEIR Package

First, call the `genSEIR` package:

```{r library}
library(genSEIR)
```

Define the country and its population:

```{r country}
country = "Italy"
Npop = 60000000 
```

Enter desired start and finish dates in *mm/dd/yy* format. Start date cannot be earlier than 01/22/21 and finish date cannot be later than the current date - 1. If `start=NULL` then start date is set to 01/22/21 and If `finish=NULL` then finish date is set to `Sys.Date()-1`:

```{r date}
start = "01/01/21" 
finish = "06/01/21"
```

Define `dt` argument which oversamples time to ensure that the algorithm converges (i.e. 1/24), and specify `f` argument for the future predictions (i.e. 30 days):  

```{r future}
dt=1
f=30
```

Get COVID-19 data from John-Hopkins University (https://github.com/CSSEGISandData/COVID-19):

```{r getData}
data = getDataCOVID(start = start, finish = finish, country = country)
```

Data for recovered cases:

```{r Recovered}
Recovered = data$tableRecovered
Recovered
```

Data for death cases:

```{r Deaths}
Deaths = data$tableDeaths
Deaths
```

Data for confirmed cases:

```{r Confirmed}
Confirmed = data$tableConfirmed
Confirmed
```

Remove first four columns since they do not include number of cases:

```{r data, results='hide'}
    recovered = Recovered[,5:ncol(data$tableRecovered)]
    deaths = Deaths[,5:ncol(data$tableDeaths)]
    confirmed = Confirmed[,5:ncol(data$tableConfirmed)]
```

Define initial estimates for parameters to be optimized: `alpha_guess`: protection rate, `beta_guess`: infection rate, `LT_guess`: latent time in days, `Q_guess`: rate at which infectious people enter in quarantine, `lambda_guess`: time dependent recovery rate, `kappa_guess`: time dependent death rate. These are the parameters that define the course of the outbreak and they will be optimized in the later steps:

```{r params}
alpha_guess = 0.2
beta_guess = 1
LT_guess = 3
Q_guess = 0.3
lambda_guess = c(0.01,0.001,10)
kappa_guess = c(0.001,0.001,10)
guess = c(alpha_guess, beta_guess, 1/LT_guess, Q_guess, lambda_guess, kappa_guess)
guess
```

Define initial conditions: `Q0`: initial number of quarantined cases, `I0`: initial number of infectious cases, `E0`: initial number of exposed cases, `R0`: initial number of recovered cases, `D0`: initial number of death cases:

```{r initial_conditions, results='hide'}
Q0 = confirmed[1]-recovered[1]-deaths[1]
I0 = 0.3*Q0
E0 = 0.3*Q0
R0 = recovered[1]
D0 = deaths[1]
```
    
Specify the number of active cases by subtracting the number of recovered and death cases from the number of confirmed cases:

```{r active_cases}
Active = confirmed-recovered-deaths
Active[Active<0] <- 0 
```

Define a time vector between start and finish dates:

```{r time_vector, results='hide'}
time = seq(as.Date(start, format = "%m/%d/%y"), as.Date(finish, format = "%m/%d/%y"), by = "1 day")
```

Optimize the outbreak parameters. The following function attempts to solve the nonlinear sum of squares problem by using Nash variant of Marquardt's approach to stabilizing the Gauss-Newton method using the Levenberg-Marquardt adjustment. This is explained in (Nash (1990) Compact numerical methods for computers: linear algebra and function minimisation. CRC press.):

```{r estimation, results='hide'}
Q=Active
R=recovered
D = deaths

params = fit_SEIQRDP(Q, R, D, Npop = Npop, E0 = E0, I0 = I0,
                         time = time, dt = dt, guess = guess, ftol = 1e-6,
                         ptol = 1e-6, gtol = 1e-6, epsfcn = 0.001, factor = 100, maxfev = 1000,
                         maxiter = 100, nprint = 1, trace = FALSE)



```

Optimized parameters:

```{r par}

params

```

Obtain predicted cases including susceptible, exposed, infectious, quarantined, recovered, dead and insusceptible using a generalized SEIR model:

```{r SEIR_results, results='hide'}

result = SEIQRDP(alpha = params$alpha1, beta = params$beta1,
              gamma = params$gamma1, delta = params$delta1,
              lambda0 = params$Lambda1, kappa0 = params$Kappa1,
              Npop, E0, I0, Q0, R0, D0, lambdaFun = params$lambdaFun,
              kappaFun = params$kappaFun, tstart = start, tfinish = finish,
              dt = dt, f = f)
    

  
```

Predictions for susceptible cases:

```{r susceptible}

head(result$susceptible)

```

Predictions for exposed cases:

```{r exposed}

head(result$exposed)

```

Predictions for infectious cases:

```{r infectious}

head(result$infectious)

```

Predictions for quarantined cases:

```{r quarantined}

head(result$quarantined)

```

Predictions for recovered cases:

```{r recovered}

head(result$recovered)

```

Predictions for dead cases:

```{r dead}

head(result$dead)

```

Predictions for insusceptible cases:

```{r insusceptible}

head(result$insusceptible)

```

Create plot for reported and predicted active, recovered and death cases. This function also plots the fitted and calculated death and recovered ratios to check whether the approximation of these ratios is appropriate:

```{r plot, results='hide'}
title =   title = paste0(country, " (Reported time range: ", start, "-", finish, "). Future predictions: ",f, " days.")

plotSEIQRDP(result, Active, recovered, deaths, title, params, checkRates = TRUE)
```


